---
title: "Descriptive Statistics and Probability in Biomedical Data Science"
author: "<br><br><span style='font-size:25px;'><strong>Lesley Chapman Hannah, Ph.D., M.S.</strong></span><br>College of Graduate Studies<br>Northeast Ohio Medical University"

format: pptx

execute:
  echo: true
  warning: false
  message: false
---

```{r, echo=FALSE}
library(tidyverse)
library(scales)
```

### Descriptive Statistics: Core Concepts

::: {style="font-size:0.65em"}
-   descriptive statistics consists of methods for organizing, summarizing, and presenting data in a way that provides insight into the characteristics of the dataset

-   involves translacting raw biological or clinical data into structured summaries and visualizations that reflect the underlying variable type and probability structure of the data

-   descriptive statistics is often the first step in understanding one's data

Before modeling, hypothesis testing, or prediction one must determine:

-   what kind of variable is being measured
-   which summaries are mathematically appropriate
-   which plots reflect the structure of the variable
:::

------------------------------------------------------------------------

### Descriptive Statistics and Variable Types

::: {style="font-size:0.65em"}
In descriptive statistics first consider:

> **What type of variable am I describing?**

-   in analysis, variables arise from different sources:

    -   mutation status \[present/absent\] → categorical
    -   tumor stage \[I–IV\] → ordinal
    -   mutation counts → discrete
    -   biomarker levels → continuous

-   each data structure determines a distinct set of statistical properties, e.x.:

-   set of allowable values the variable can take

-   how variability appears across observations

-   which numerical summaries appropriately represent the data

-   which visualizations accurately reflect the distribution

-   starting with continuing the variable being described you allow for:

    -   summaries align with the data-generating process
    -   visualizations reflect the true distribution of values
:::

------------------------------------------------------------------------

### Variable Types Used in Descriptive Statistics

Variable types used in analysis:

-   **categorical nominal variables**
-   **ordinal variables**
-   **count variables**
-   **continuous variables**

------------------------------------------------------------------------

### How Probability Connects to Description

-   probability tells us how often values or outcomes occur

-   probability defined based on outcome variables:

    -   categorical variables: probabilities of categories
    -   ordinal variables: probabilities across ordered levels
    -   count variables: probabilities of integer counts
    -   continuous variables: density across a numeric range

------------------------------------------------------------------------

### Code Example Primer

**Goal**: highlight how descriptive statistics can be used as summary and how plots are chosen based on variable type

1.  generate a sample from a hypothetical population\
2.  identify four different variable types\
3.  compute probability-oriented summaries for each type\
4.  match each variable to an appropriate plot\
5.  interpret how variable structure determines description

------------------------------------------------------------------------

### Code Example

::: {style="font-size:0.7em"}
**Background**: you received data from a pilot study designed to characterize inflammatory signaling and genomic alteration patterns in a cohort of 160 patients

Each patient represents a single observational unit sampled from a broader clinical population. The goal is to describe how molecular and clinical features vary across individuals and how these features are structured in the data.

For each patient, we record:

**mutation status**: mutation absent or present \[categorical nominal\]

**tumor stage**: ordered clinical classification that describes how far a cancer has progressed in the body - I,II,III,IV \[ordinal\]

**mutation count** from a targeted sequencing panel: count of the number of times an alleles at a genomic locus - 0,1,2 \[count\]

**IL-6 concentration \[pg/mL\]**: as a circulating inflammatory biomarker \[continuous\]
:::

------------------------------------------------------------------------

### Code Example

::: {style="font-size:0.7em"}
```{r, echo=FALSE}
set.seed(301)

biomed_sim <- tibble(
  patient_id = sprintf("P%03d", 1:160),
  mutation_status = sample(
    c("Absent", "Present"),
    size = 160,
    replace = TRUE,
    prob = c(0.68, 0.32)
  ),
  tumor_stage = sample(
    c("I", "II", "III", "IV"),
    size = 160,
    replace = TRUE,
    prob = c(0.22, 0.33, 0.27, 0.18)
  ),
  mutation_count = rpois(160, lambda = 3.1),
  il6_pg_ml = round(rlnorm(160, meanlog = log(7.5), sdlog = 0.55), 2)
) |>
  mutate(
    mutation_status = factor(mutation_status),
    tumor_stage = factor(tumor_stage, levels = c("I", "II", "III", "IV"), ordered = TRUE)
  )
```

```{r}
biomed_sim |> slice_head(n = 10)
```

**Key concepts**

**Why this simulation supports the lecture concepts**

-   mutation status shows a **categorical probability distribution**
-   tumor stage shows an **ordered probability distribution**
-   mutation count shows a **discrete count distribution**
-   IL-6 shows a **continuous distribution** across a numeric scale
:::

------------------------------------------------------------------------

### Code Example: Inflammatory Signaling Pathway

::: {style="font-size:0.6em"}
```{r, echo=FALSE}
library(tibble)
library(knitr)

tbl <- tibble(
  variable = c("mutation_status", "tumor_stage", "mutation_count", "il6_pg_ml"),
  type = c("Categorical nominal", "Ordinal", "Count", "Continuous"),
  values = c(
    "Absent / Present",
    "I < II < III < IV",
    "0, 1, 2, ...",
    "Positive real values"
  ),
  descriptive_goal = c(
    "Counts and proportions by category",
    "Counts and proportions across ordered levels",
    "Frequency of each integer count",
    "Center, spread, and distribution shape"
  ),
  plot = c(
    "Bar plot",
    "Ordered bar plot",
    "Bar plot of count frequencies",
    "Histogram"
  )
)
```

```{r}
kable(tbl, align = "lllll")
```
:::

------------------------------------------------------------------------

### Code Example: Inflammatory Signaling Pathway

::: {style="font-size:0.6em"}
```{r}
mutation_summary <- biomed_sim |>
  count(mutation_status) |>
  mutate(prob = n / sum(n))

mutation_summary
```

**Key parts of the code**

-   `count(mutation_status)` counts the number of patients in each category
-   `mutate(prob = n / sum(n))` converts counts into proportions
-   `prob` column: probability distribution for mutation status in this sample

**How to interpret the output**

-   `n` gives the observed number of patients in each class
-   `prob` gives the estimated probability that a randomly selected patient from this sample falls into that class
-   probabilities sum to 1 because every patient must belong to exactly one mutation-status category
:::

------------------------------------------------------------------------

### Categorical Example: Plot

::: {style="font-size:0.6em"}
```{r}
ggplot(mutation_summary, aes(x = mutation_status, y = prob)) +
  geom_col() +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Mutation Status: Empirical Category Probabilities",
    x = "Mutation status",
    y = "Sample proportion"
  ) + theme_bw()
```

**Interpretation**

-   categorical variables are described by **counts, proportions, and category probabilities**
-   X-axis: contains discrete categories
-   bar height: represents the probability associated with each category
:::

------------------------------------------------------------------------

### Ordinal Example

::: {style="font-size:0.6em"}
```{r}
stage_summary <- biomed_sim |>
  count(tumor_stage) |>
  mutate(prob = n / sum(n))

stage_summary
```

**Code Overview**

-   `tumor_stage`: coded as an ordered factor
-   `count()`: summary of the number of samples within each tumor stage
-   `prob = n / sum(n)`: converts counts into sample probabilities

**Notes**

-   stage I through IV define a progression
-   preserving order makes the summary more interpretable
:::

------------------------------------------------------------------------

### Ordinal Example

::: {style="font-size:0.6em"}
```{r}
ggplot(stage_summary, aes(x = tumor_stage, y = prob)) +
  geom_col() +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Tumor Stage: Ordered Distribution Across Levels",
    x = "Tumor stage",
    y = "Sample proportion"
  ) + theme_bw()
```

**Interpretation**

-   bar plot chosen to show variable has discrete levels
-   levels are arranged in clinical order from I to IV
-   nominal categorical variable vs. ordinal variable: both use proportions, but the ordinal variable includes rank information \[order of bars from stage 1 $\rightarrow$\]
:::

------------------------------------------------------------------------

### Mutation Count Example

::: {style="font-size:0.6em"}
```{r}
library(dplyr)
library(tidyr)

count_summary <- biomed_sim |>
  count(mutation_count) |>
  complete(mutation_count = 0:2, fill = list(n = 0)) |>
  arrange(mutation_count) |>
  mutate(prob = n / sum(n))

count_summary
```

**Code Overview**

`count(mutation_count)`: counts observed frequencies for each mutation count `complete(mutation_count = 0:2, fill = list(n = 0))`: mutation counts at {0,1,2} `prob = n / sum(n)`: converts counts into a probability mass function (PMF)

**Interpretation**:

`mutation_count = 0`: probability that no mutations are observed `mutation_count = 1`: probability of exactly one mutation `mutation_count = 2`: probability of exactly two mutations
:::

------------------------------------------------------------------------

### Mutation Count Example

::: {style="font-size:0.6em"}
```{r, echo=FALSE}
p_count <- ggplot(count_summary, aes(x = factor(mutation_count), y = prob)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Mutation Count: Empirical Probability Mass Function",
    x = "Mutation count",
    y = "Sample proportion"
  ) +
  theme_bw()

p_count
```

**Interpretation**

-   each bar represents the probability of observing a specific mutation count within the defined support {0,1,2}
-   mutation count is restricted to a finite set of allowable integer values {0,1,2} $\rightarrow$ x-axis represents distinct, unordered count levels
-   height of each bar corresponds to the probability P(X=x) for a specific count
-   bars are separated because probabilities are assigned to individual count outcomes
:::

------------------------------------------------------------------------

### Continuous Example: IL-6

::: {style="font-size:0.6em"}
```{r}
il6_summary <- biomed_sim |>
  summarise(
    n = n(),
    mean_il6 = mean(il6_pg_ml),
    median_il6 = median(il6_pg_ml),
    sd_il6 = sd(il6_pg_ml),
    iqr_il6 = IQR(il6_pg_ml),
    min_il6 = min(il6_pg_ml),
    max_il6 = max(il6_pg_ml)
  )

il6_summary
```

**Key parts of the code**

-   `summarise()` reduces the full vector of IL-6 values into a set of descriptive statistics
-   `mean()` and `median()`: describe summary statistics
-   `sd()` and `IQR()`: data spread - standard deviation and interquartile range
-   `min()` and `max()` : defines the observed range

**Interpretation**

-   if the mean \> median: data is often right skewed
-   larger standard deviation : greater dispersion around the mean
-   IQR captures the width of the middle 50% of the data
:::

------------------------------------------------------------------------

### Continuous Example: IL-6

::: {style="font-size:0.6em"}
```{r, echo=FALSE}
# compute quartiles
q1 <- quantile(biomed_sim$il6_pg_ml, 0.25)
q3 <- quantile(biomed_sim$il6_pg_ml, 0.75)
med <- median(biomed_sim$il6_pg_ml)

plot2 <- ggplot(biomed_sim, aes(x = il6_pg_ml)) +
  geom_histogram(bins = 20) +
  
  # IQR shaded region
  annotate("rect",
           xmin = q1, xmax = q3,
           ymin = 0, ymax = Inf,
           alpha = 0.2) +
  
  # median line
  geom_vline(xintercept = med, linetype = "dashed") +
  
  labs(
    title = "IL-6 Distribution Across the Simulated Sample",
    subtitle = "Shaded region = IQR (middle 50%), dashed line = median",
    x = "IL-6 (pg/mL)",
    y = "Number of patients"
  ) +
  theme_bw()

plot2
```

**Interpretation**

-   IL-6 is measured on a continuous numeric scale
-   histogram groups nearby values into bins → overall distributional shape becomes visible
-   shaded region represents the interquartile range (IQR), capturing the middle 50% of observed values
    -   IQR spans from the 25th percentile (Q1) to the 75th percentile (Q3)
-   dashed line marks the median, $\rightarrow$ central location of the distribution
:::

------------------------------------------------------------------------

### Summary

::: {style="font-size:0.62em"}
-   analysis begins by identifying what kind of variable is being measured in descriptive statistics

-   review for descriptive statistics:

observed values → variable type → probability structure → summary → plot observed values

-   `categorical variables` $\rightarrow$ described with counts, proportions, and bar plots

-   `ordinal variables` $\rightarrow$ described with ordered counts, ordered proportions, and plots that preserve rank

-   `count variables` $\rightarrow$ described through frequencies of integer outcomes and probability mass functions

-   `continuous variables` $\rightarrow$ described through center, spread, and distribution shape using summaries such as the mean, median, SD, and IQR

    -   plots: histograms or density-based displays

**Next Lecture**: more focused discussion on variable identification and determining the descriptive metric, probability summary, and plot
:::

------------------------------------------------------------------------
